---
---
title: "Assignment 1: Setting Up R"
author: "Robert Andoh"
date: "`r Sys.Date()`"
output:
  HTML_document: default
---

# Question 1

Installed the ISLR package and loaded the library.

```{r}
library(ISLR)
```

# Question 2

Print summary statistics and count rows.

```{r}
summary(Carseats)
nrow(Carseats)
```

Answer:

The Carseats dataset contains **400 observations**.

# Question 3

```{r}
max(Carseats$Advertising)
```

Answer:

The maximum Advertising value is **29**.

# Question 4

```{r}
IQR(Carseats$Price)
```

Answer:

The IQR of Price is **31**.

# Question 5

```{r}
plot(Carseats$Price,
     Carseats$Sales,
     main="Sales vs Price",
     xlab="Price",
     ylab="Sales")

cor(Carseats$Price,
    Carseats$Sales)
```

Answer:

The scatterplot shows a negative relationship between Sales and Price.

The correlation coefficient is approximately **-0.445**, indicating a moderate negative relationship.

# Question 6

This notebook contains:
- A title
- Descriptions
- Comments
- R code
- Results
