---
title: "Assignment 2: k-NN Classification for Universal Bank Personal Loan Prediction"
author: "Robert Andoh"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

# Introduction

This assignment applies the k-Nearest Neighbors (k-NN) classification algorithm to predict whether a Universal Bank customer will accept a personal loan offer. The analysis includes data preparation, model building, model evaluation, customer classification, and performance comparison using different data partitions.

# Load Required Packages

The following packages are required for data partitioning, model building, and model evaluation.

```{r}
library(class)
library(caret)
```

# Import and Inspect the Dataset

The Universal Bank dataset is imported into R and examined to understand its structure and dimensions.

```{r}
bank <- read.csv("UniversalBank-1.csv")

head(bank)
dim(bank)
str(bank)
```

# Distribution of Personal Loan Variable

The target variable is Personal.Loan where:

- 0 = Customer did not accept the loan offer
- 1 = Customer accepted the loan offer
```{r}
table(bank$Personal.Loan)

prop.table(table(bank$Personal.Loan))
```